Exploring probability distributions for bivariate temporal granularities

### 2019-12-02 main.Rmd
synergy includec in data structure and cyclic intro change

Periodicity is very common in all kinds of data. Time granularities which can accomodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. Hence, time granularities can also be **circular** or **aperiodic** depending on if the mapping of the linear granularities is regular or irregular. Mappings will be regular if, for example, they are formed by grouping operation. Grouping operation would ensure that each granule of the resulting granularity consists of same number of granules of bottom granularity. However, if they are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of bottom granularity. However, the formalism would differ for circular (regular mapping) and quasi-circular (irregular mapping) granularities. Examples of circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or public holidays.

@@

# Synergy of cyclic time granularities {#sec:synergy}

@@

Before exploring the behavior of a "dependent variable" across time granularities,
it is important to know how these granularities interact with each other. If two time granulaties of interest interact, the relationship between each of the interacting variables and the dependent variable will depend on the value of the other interacting variable. To define a general framework, we define harmony and clashes.

@@

Suppose we have two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1,A_2,A_3,\dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1,B_2,B_3,\dots, B_m$\}.  That is, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = i$ and $C_2(s) = j$. Since, $C_1$ has n levels and $C_2$ has m levels there will be $nm$ such sets $S_{ij}$.Now, many situations can lead to any of these sets being empty. Let us discuss the following cases, where one or more of these sets can be empty.

@@

Therefore, pair of circular/aperiodic granularities which lead to structurally, event-driven or build-based empty-combinations are referred to as to as **clashes**.
And the ones that do not lead to any missing combinations are called **harmonies**.

Creating good visualization requires good data structures. Commonly, simple sequences of time value pairs $<t_0, v_0> \dots, <t_n, v_n>$ are the basis of analysis and visualization. It is recognized that the initial approaches of just considering time as an ordinal dimension in a 2D or 3D visualisation are inadequate to capture characteristics of time-dependent information. For example, there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use different calendars that define meaningful system of granularities for different application domain. Data can then be integrated at different levels of granularity enabling value aggregation or distribution along granularities. Since, we are essentially concerned with detection of previously unknown periodic behavior of data, it makes sense to support the detection of patterns by anabling aggregation or distribution across cyclic time granularities.

 Instead, the big problem can be broken down into smaller pieces by focusing on specific tasks per visual representation in a 2D space.

>>>>

Periodicity is very common in all kinds of data. Time granularities which can accomodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. Cyclic time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the same is irregular. Examples of circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or public holidays. To elaborate more, mappings will be regular if, for example, linear granularities are formed by grouping operation. Grouping operations combine fixed granules of a linear granularity to form a new granule of the resulting linear granularity. However, if granularities are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of the root granularity. Both circular and quasi-circular granularities accomodate periodicities in data. However, the formalisms would differ.

@@

Effective exploration or good visualization require good data structures. Temporal data problems are typically grouped into two types of analysis, time series and longitudinal. A new tidy data structure to support exploration and modeling of temporal data is tsibble, where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. This opens the door to easily operating on time to make calculations and extract different temporal components, such as month and day of the week. Commonly, simple sequences of time value pairs $<t_0, v_0> \dots, <t_n, v_n>$ are the basis of analysis and visualization. It is recognized that the initial approaches of just considering time as an ordinal dimension in a 2D or 3D visualisation are inadequate to capture characteristics of time-dependent information. For example, there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use different calendars that define meaningful system of granularities for different application domain. Data can then be integrated at different levels of granularity enabling value aggregation or distribution along granularities. Since, we are essentially concerned with detection of previously unknown periodic behavior of data, it makes sense to support the detection of patterns by anabling aggregation or distribution across cyclic time granularities. We consider the following data structure for exploring distribution of a measured variable ((Figure \@ref(fig:data-structure)). These are based on conditioning the values taken on by one or more of the cyclic granularities in a data set. This means data subsets are formed for each level of the conditioned variable and the fixed variable. This can be an useful data structure for exploratory data analysis as we can rapidly compare patterns in different parts of the data.

```{r data-structure, fig.cap = "The data structure for exploring periodicities along cyclic granularities. Two cyclic granularities serve as fixed and conditioned variable and the observational unit across combinations of these cyclic time granularities.", fig.pos="htb"}
include_graphics("Figs/data-structure.png")
```

## Fixed and Conditioned variables {#sec:synergy}

The cyclic time granularities which are essentially a function of the index of the tsibble are abstractions of time and hence are discrete variable, mapped as a factor. The relationships between the fixed and conditioned variable are of importance in this data structure due to the ways the cyclic granularities, which are mapped to fixed and conditional variable, relate to one another. To define a general framework, we define harmony and clashes.

@@

<!-- Before exploring the behavior of a "dependent variable" across time granularities, -->
<!-- it is important to know how these granularities interact with each other. If two time granulaties of interest interact, the relationship between each of the interacting variables and the dependent variable will depend on the value of the other interacting variable.  -->

@@

Suppose we have two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1,A_2,A_3,\dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1,B_2,B_3,\dots, B_m$\}.  That is, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = i$ and $C_2(s) = j$. Since, $C_1$ has n levels and $C_2$ has m levels there will be $nm$ such sets $S_{ij}$. Now, many situations can lead to any of these sets being empty. Let us discuss the following cases, where one or more of these sets can be empty.

@@

Therefore, pair of cyclic granularities which lead to structurally, event-driven or build-based empty-combinations are referred to as to as **clashes**. And the ones that do not lead to any missing combinations are called **harmonies**.

## Measured variable

The measured variable which is the univariate observational units can be dicrete or continuous.

<!-- would invlove a faceting approach with one temporal granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. -->

 <!-- Instead, the big problem can be broken down into smaller pieces by focusing on specific tasks per visual representation in a 2D space. -->
